---
title: "Get study effort"
output: 
  github_document:
    toc: true
    df_print: kable
---

I want to assess taxonomic bias in the recovery rate dataset. Are species in that dataset more intensely studied? 

```{r setup, include=FALSE}
library(tidyverse)
library(rentrez)
library(ape)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
theme_set(new = theme_bw())
```

I import three datasets: (i) open tree taxonomy for acanths, cestodes, and nematodes, (ii) life cycle database, and (iii) recovery rate dataset. We want to extract all genera from these data and then download the study effort. In this way, we can compare whether the groups make up some datasets have been more intensely studied than others.

```{r importdata}
# import LCDB
dat_host <- read.csv(file = "../../data/CLC_database_hosts.csv", header = TRUE)
dat_lh <- read.csv(file = "../../data/CLC_database_lifehistory.csv", header = TRUE)
# import recovery rate data
dat <- read.csv(file = "../../data/er_combined.csv", header = T)
# import taxonomy
ac <- read.csv(file = "../../data/acanth_taxonomy.csv", header = T)
ce <- read.csv(file = "../../data/cest_taxonomy.csv", header = T)
ne <- read.csv(file = "../../data/nem_taxonomy_rotl.csv", header = T)
```
```{r}
# combine datasets, retaining species, worm group, and data source
spp <- bind_rows(
  select(dat_host, Parasite.species, Parasite.group)%>%mutate(db = "lcdb"),
  select(dat_lh, Parasite.species, Parasite.group)%>%mutate(db = "lcdb"),
  select(dat, Parasite.species = Species, Parasite_Group)%>%
    mutate(Parasite.group = tolower(Parasite_Group), db = "rr"),
  select(ac, Parasite.species = species)%>%mutate(Parasite.group = "acanthocephalan", db = "ot"),
  select(ce, Parasite.species = species)%>%mutate(Parasite.group = "cestode", db = "ot"),
  select(ne, Parasite.species = species)%>%mutate(Parasite.group = "nematode", db = "ot")
)%>%
  select(Parasite.species, Parasite.group, db)%>%
  distinct()

spp <- spp%>%
  mutate(Parasite.genus = substr(Parasite.species, start = 1, stop = regexpr(" ", Parasite.species)-1))
```

```{r}
# make genus-level db
g_lev <- spp%>%
  filter(Parasite.genus != "")%>%
  group_by(Parasite.genus, Parasite.group, db)%>%
  summarise(n_spp = n())%>%
  ungroup()%>%
  pivot_wider(names_from = "db", values_from = n_spp, names_prefix = "n_spp_")%>%
  mutate(across(starts_with("n_spp_"), function(x){ifelse(is.na(x), 0, x)}))
```

For every unique genus in the data, we query pubmed and extract the number of publication "hits". This is a measure of study effort. As a control, we compare this to the number of studies for a genus in the life cycle database.

```{r}
# Number of unique publications in life cycle database
dat_host <- mutate(dat_host, publication = paste(Author, Year, Journal, Volume, Pages)) # creat publication variable
dat_lh <- mutate(dat_lh, publication = paste(Author, Year, Journal, Volume, Pages))

dat_hosts_pub <- filter(dat_host, !is.na(Author), !is.na(Year), !is.na(Journal))%>%
  select(Parasite.genus, publication)%>%
  distinct()
dat_lh_pub <- filter(dat_lh, !is.na(Author), !is.na(Year), !is.na(Journal))%>%
  select(Parasite.genus, publication)%>%
  distinct()

study_effort <- bind_rows(dat_hosts_pub, dat_lh_pub)%>%arrange(Parasite.genus, publication)%>%distinct()

study_effort <- group_by(study_effort, Parasite.genus)%>%
  summarise(n_pubs_lcdb = n())
```

```{r}
g_lev <- left_join(g_lev, study_effort)
```

For querying PubMed, we create Entrez queries like so: "parasite genus[ALL] AND parasite group[All]". We include parasite group so that the query does not return hits on non-helminth groups.

```{r}
# make queries for genus name plus parasite group
g_lev <- g_lev%>%
  mutate(p_g_wc = if_else(Parasite.group == 'nematode', 'nematod*', 
                          if_else(Parasite.group == 'cestode', 'cestod*', 'acanthocephala*')))%>%
  mutate(query = paste0(Parasite.genus, '[ALL] AND ', p_g_wc, '[ALL]'))%>%
  select(-p_g_wc)
query_gen_name_group <- g_lev$query
```

```{r}
for(query in query_gen_name_group){
  # search
  search_res <- entrez_search(db = "pubmed", term = query)
  # report hits
  num_hits <- search_res$count
  print(paste(num_hits, "hits for", query))
  
  if(query == query_gen_name_group[1]){
    pubmed_hits2 <- num_hits
  } else {
    pubmed_hits2 <- c(pubmed_hits2, num_hits)
  }
}

```

```{r}
g_lev <- left_join(g_lev, data.frame(query = query_gen_name_group,
                              pubmed_hits = pubmed_hits2)
)

```

Here is the distribution of pubmed hits based on genus name + parasite group. It is skewed, even on a log scale

```{r}
ggplot(g_lev, aes(x = pubmed_hits+1)) + geom_histogram() + scale_x_log10()
```

Are the number of pubmed hits correlated with the number of publications on a given genera in the life cycle database? It is, but it is noisy, which is not too surprising, as I was not trying to be exhaustive when compiling the database.

```{r}
ggplot(g_lev, aes(x = n_pubs_lcdb+1, y = pubmed_hits+1)) + 
  geom_point(alpha = 0.2) + 
  geom_smooth() +
  scale_y_log10() + scale_x_log10()
```

```{r}
write.csv(g_lev, file = "../../data/study_effort_genus.csv", row.names = F)
```

